Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2023-10-27 sections/combinatorial_semantics.tex
Draft of pushout for EHyp

    The pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X} + s_{Y}, t_{X} + t_{Y}, \#_{X} + \#_{Y}, <_{c,X} + <_{c,Y} \}
    once again omitting the labelling function.
    And consider a relation $S = \{(x,y), (y,x) \in (X+Y) \times (X+Y) \; | \;\exists z \in Z \; . \; f(z) = x \text{ and } g(z) = y \}$ and let $R$ be its reflexive and transitive closure.
    We then quotient $X + Y$ by $R$ and claim that it is a pushout.
        \sfrac{X + Y}{\sim R} = \{\sfrac{V_{X} + V_{Y}}{\sim R}, \sfrac{E_{X} + E_{Y}}{\sim R}, \sfrac{s_{X} + s_{Y}}{\sim R}, \sfrac{t_{X} + t_{Y}}{\sim R}, \sfrac{\#_{X} + \#_{Y}}{\sim R}, \sfrac{<_{c,X} + <_{c,Y}}{\sim R} \}    
    where quotienting on source and target functions is taken as in $Hyp_{\Sigma}$.
    \sfrac{(<_{c,X+Y})}{\sim (R_{V},R_{E})} \subseteq \sfrac{(V_{X} + V_{Y})}{\sim R_{V}} + \sfrac{(E_{X} + E_{Y})}{\sim R_{E}} \times \sfrac{(E_{X} + E_{Y})}{\sim R_{E}}
    \update{We shall probably first define the quotient of the immediate parent because it is essentially a function and it will follow the same argument as the source function}
    \question{Below $e$ and $v$ do not necessarily mean edges and nodes, $v$ can be an edge as well.}
    \question{I also kind of intuitively inventing the quotient of a relation, is there any accepted notion for this?}
    and, whenever $e_{X} <_{c,X}^{\mu} v_{X}$ and $e_{Y} <_{c,Y}^{\mu} v_{Y}$ and $v_{x} R v_{y}$ then $e_{X} R e_{Y}$.
    The latter condition means two things
        \label{cond:essential_for_pushout}
        \item if $[z) = \varnothing$ then either $[f(z)) = \varnothing$ or $[g(z)) = \varnothing$.
              \question{This point formally defines what it means to 'inherit' a relation.
                        Let's say $v_1 = f(z)$ and $v_2 = g(z)$ and $e_1 <_{c,X}^{\mu} v_1$ and $[v_2) = \varnothing$,
                        then, having $e_1 <_{c} [v_1,v_2]$ is fine as it does not break the condition of a predecessor being unique.}
              \update{This point also says when we can compute a pushout: either non of X and Y should have a parent for $f(z)$ and $g(z)$ or only one of them should.}
        \item if $[z) \not = \varnothing$ then there exists $z'$ such that $z' <_{c}^{\mu} z$ and $[g(z'), f(z')] <_{c}^{\mu} [g(z), f(z)]$
        \sfrac{\conflict_{X} + \conflict_{Y}}{\sim R} \subseteq \sfrac{(V_{X} + V_{Y} + E_{X} + E_{Y})}{\sim R} \times \sfrac{(E_{X} + E_{Y})}{\sim R} \times \sfrac{(V_{X} + V_{Y} + E_{X} + E_{Y})}{\sim R} \times \sfrac{(V_{X} + V_{Y} + E_{X} + E_{Y})}{\sim R}
    meaning that whenever $e_1 \conflict e_2$ and $e_1' \conflict e_2'$ and $e_1 R e_1'$ then $[e_1, e_1'] \conflict e_2$ and $[e_1, e_1'] \conflict e_2'$.
    $\sfrac{i_1}{\sim R}$ then maps $x \to [x]$ and $\sfrac{i_2}{\sim R}$ maps $y \to [y]$.
    Before we proceed we need to check that $\sfrac{(X + Y)}{\sim R}$ is indeed an e-hypergraph.
    Since $X + Y$ is clearly an e-hypergraph, we need to check whether identification breaks any of the requirements (1-7) above.
        \item Assume $[x_1,y_1] <_{c} [x_2,y_2]$ and $[x_1, y_1] \conflict [x_3]$ which means that $x_1 \conflict x_3$.
              Hence $x_2 \conflict x_3$ (because $X$ is an e-hypergraph) and $[x_2,y_2] \conflict [x_3]$. Similarly, if $[x_1, y_1] \conflict [y_3]$ and $y_1 \conflict y_3$.
        \item $\sfrac{\conflict}{\sim R}$ and $\sfrac{<_{c}}{\sim R}$ are disjoint because we quotient only inside the relations.
        \item Again, we only quotient $\conflict$ on $x$ and $y$ for which original relation is defined.
        \item This follows from $\sfrac{i_1}{\sim R}$ and $\sfrac{i_2}{\sim R}$ being homomorphisms.
        \item We just perform transitive closure of the complement. \update{Maybe we can get rid of this point and instead later just take transitive closure of the complement when talking about interfaces}
        \item Assume $[x_1, y_1] \conflict [x_2,y_2]$ and $[x_2,y_2] \conflict [x_3,y_3]$ and all of them are distinct.
              Then $[x_1,y_1] \conflict [x_3,y_3]$ because either $x_1 \conflict x_3$ or $y_1 \conflict y_3$ (otherwise they would not have been merged into the same equivalence class).
        \item We never merge $x_1$ and $x_2$ such that $x_1 <_{c} x_2$ into the same equivalence class because morphisms should preserve the order.

>>>>

    The pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \#_{X+Y}, <_{c,X+Y} \}
    \] 
    once again omitting the labelling function and where source and target functions are as before.
    And consider the same relations $S_{V},S_{E}$ and $R_{V}, R_{E}$.
    We then quotient $X + Y$ by $R_{V},R_{E}$ and claim that it is a pushout.
    \[
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{\sfrac{V_{X} + V_{Y}}{\sim R_{V}}, \sfrac{E_{X} + E_{Y}}{\sim R_{E}}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})}, \sfrac{t_{X,Y}}{\sim ()}, \sfrac{\#_{X+Y}}{\sim (R_{V},R_{E})}, \sfrac{<_{c,X + Y}}{\sim (R_{V},R_{E})} \}    
    \]
    where quotienting on source and target functions is taken as in $Hyp_{\Sigma}$ and we will further omit writing the relations explicitly.
    \[
    \sfrac{(<_{c,X+Y})}{\sim} \subseteq \sfrac{(V_{X} + V_{Y})}{\sim R_{V}} + \sfrac{(E_{X} + E_{Y})}{\sim R_{E}} \times \sfrac{(E_{X} + E_{Y})}{\sim R_{E}}
    \]
    such that for each $(x_1, x_2) \in <_{c, X+Y}$ $\sfrac{<_{c,X+Y}}{\sim}$ contains $([x_1],[x_2])$.

    \update{We shall probably first define the quotient of the immediate parent because it is essentially a function and it will follow the same argument as the source function}
    \question{Below $e$ and $v$ do not necessarily mean edges and nodes, $v$ can be an edge as well.}
    \question{I also kind of intuitively inventing the quotient of a relation, is there any accepted notion for this?}
    and, whenever $e_{X} <_{c,X}^{\mu} v_{X}$ and $e_{Y} <_{c,Y}^{\mu} v_{Y}$ and $v_{x} R v_{y}$ then $e_{X} R e_{Y}$.
    The latter condition means two things
    \begin{itemize}
        \label{cond:essential_for_pushout}
        \item if $[z) = \varnothing$ then either $[f(z)) = \varnothing$ or $[g(z)) = \varnothing$.
              \question{This point formally defines what it means to 'inherit' a relation.
                        Let's say $v_1 = f(z)$ and $v_2 = g(z)$ and $e_1 <_{c,X}^{\mu} v_1$ and $[v_2) = \varnothing$,
                        then, having $e_1 <_{c} [v_1,v_2]$ is fine as it does not break the condition of a predecessor being unique.}
              \update{This point also says when we can compute a pushout: either non of X and Y should have a parent for $f(z)$ and $g(z)$ or only one of them should.}
        \item if $[z) \not = \varnothing$ then there exists $z'$ such that $z' <_{c}^{\mu} z$ and $[g(z'), f(z')] <_{c}^{\mu} [g(z), f(z)]$
    \end{itemize}
    Then,
    \[
        \sfrac{\conflict_{X} + \conflict_{Y}}{\sim R} \subseteq \sfrac{(V_{X} + V_{Y} + E_{X} + E_{Y})}{\sim R} \times \sfrac{(E_{X} + E_{Y})}{\sim R} \times \sfrac{(V_{X} + V_{Y} + E_{X} + E_{Y})}{\sim R} \times \sfrac{(V_{X} + V_{Y} + E_{X} + E_{Y})}{\sim R}
    \]
    meaning that whenever $e_1 \conflict e_2$ and $e_1' \conflict e_2'$ and $e_1 R e_1'$ then $[e_1, e_1'] \conflict e_2$ and $[e_1, e_1'] \conflict e_2'$.
    $\sfrac{i_1}{\sim R}$ then maps $x \to [x]$ and $\sfrac{i_2}{\sim R}$ maps $y \to [y]$.
    Before we proceed we need to check that $\sfrac{(X + Y)}{\sim R}$ is indeed an e-hypergraph.
    Since $X + Y$ is clearly an e-hypergraph, we need to check whether identification breaks any of the requirements (1-7) above.
    \begin{enumerate}
        \item Assume $[x_1,y_1] <_{c} [x_2,y_2]$ and $[x_1, y_1] \conflict [x_3]$ which means that $x_1 \conflict x_3$.
              Hence $x_2 \conflict x_3$ (because $X$ is an e-hypergraph) and $[x_2,y_2] \conflict [x_3]$. Similarly, if $[x_1, y_1] \conflict [y_3]$ and $y_1 \conflict y_3$.
        \item $\sfrac{\conflict}{\sim R}$ and $\sfrac{<_{c}}{\sim R}$ are disjoint because we quotient only inside the relations.
        \item Again, we only quotient $\conflict$ on $x$ and $y$ for which original relation is defined.
        \item This follows from $\sfrac{i_1}{\sim R}$ and $\sfrac{i_2}{\sim R}$ being homomorphisms.
        \item We just perform transitive closure of the complement. \update{Maybe we can get rid of this point and instead later just take transitive closure of the complement when talking about interfaces}
        \item Assume $[x_1, y_1] \conflict [x_2,y_2]$ and $[x_2,y_2] \conflict [x_3,y_3]$ and all of them are distinct.
              Then $[x_1,y_1] \conflict [x_3,y_3]$ because either $x_1 \conflict x_3$ or $y_1 \conflict y_3$ (otherwise they would not have been merged into the same equivalence class).
        \item We never merge $x_1$ and $x_2$ such that $x_1 <_{c} x_2$ into the same equivalence class because morphisms should preserve the order.
    \end{enumerate}
